---
title: "Assignment 4"
author: "Dakota Hawkins & Mae Rose Gott"
date: "2/14/2022"
output: html_document
---

# Welcome to Assignment 4

## Before we begin

Unlike Assignments 1-3, you will no longer be provided with the RMarkdown code.
All of the code blocks here are empty and accompanied by instructions for what is 
expected in the code block. Do not neglect your `main.R` skeleton file as the 
functions there will constitute most of the points you earn. 

_THE BULK OF YOUR GRADE WILL BE ON YOUR WORK IN `main.R`_

Anything that is specified to occur in functions you create MUST be in the `main.R`
file. While you theoretically CAN do some or all of the homework in this .Rmd file, 
it is bad practice to have your 'back end' coding in the same files as your 'front
end' report. Your `report.html` may be referenced to visually check any graphical
outputs you may generate for this and all future assignments, but your `main.R` 
functions will be run outside of their respective markdown environments.

```{r}
#Source your main.R

```

## 1.  Reading and filtering count data

To begin, let's read the raw count matrix, named `verse_counts.tsv`, and filter out all zero-variance genes. Filtering out zero-variance genes is necessary for several common analysis steps, such as Principal Component Analysis (PCA), and can be helpful to reduce the size of our dataset if memory issues are a constraint. Implement the `read_data` function in `main.R` to read in the dataset, then use the `filter_zero_var_genes` function, also in `main.R` to remove zero variance genes from the count matrix, and return a list containing both the name of the returned genes, and the counts of each gene for every sample.

1a. import data
```{r}
#Read in verse_counts.tsv and display the head of the resulting tibble

```

1b. Apply zero variance filter
```{r}
#Filter out zero variance genes using the function you created in `main.R` and 
#display the first 9 genes from the zero variance filter function's output

```

1c. Display the counts matrix
```{r}
#display the head of the counts matrix from the zero variance filter function's output

```

## 2. Constructing sample meta data from sample names
Often during analysis, it's beneficial to separate data modes from each other (
i.e. to have one data frame containing gene expression data, and separate data
frame for any gene or subject level information). Here we'll construct a `tibble`
from sample names located in the columns of our count matrix. Implement the 
`meta_info_from_labels` function to extract the sample name, timepoint, and
replicate number from each sample name. The function should then return a
`tibble` with all the relevant information.
```{r}
#Create and display the sample meta data tibble

```

## 3. Count distributions and normalization
Often we will want to normalize gene counts between samples so that comparisons
are more reliable. Given read counts are a relative measure of abundance,
normalization is a necessary step during RNAseq analysis when samples may have
differing numbers of total reads. 

### 3a. Visualizing count distributions via boxplots 
To visualize these differences, implement the
`plot_sample_distributions()` function using `ggplot2` to create boxplots 
visualizing the read counts for each sample over all genes. Ensure the function
is able to scale the y-axis to $log$ values if specified by the `scale_y_axis`
parameter. Color each boxplot by its respective sample name.

*Hint*: make sure your `plot_sample_distributions()` reshapes your count matrix 
from a _wide_ format, to a _long_ format. 
```{r}
#Display your count distributions boxplot

```

### 3b. Count variance vs mean count
Another aspect of RNAseq data that often needs to be accounted for the is 
relationship between variance and mean expression: that is, more highly
expressed genes will exhibit higher variance values. This is problematic as
interesting patterns controlled by genes at naturally lower expression values
will be washed out by more highly expressed genes. To visualize this phenomenon,
implement the `plot_variance_vs_mean()` function to plot gene variance on the
y-axis vs mean count rank on the x-axis, where with $p$ genes, the most highly
expressed gene will have rank $p$ and the most lowly expressed gene will have
rank $1$. 
```{r}
#Display your variance vs mean plot

```


### 3c. Visualizing patterns with PCA
Principal Component Analysis (PCA) is a dimension reduction technique often used
to extract the most meaningful structure in a dataset to fewer dimensions. This
is incredibly useful when trying to visualize your data in only a few dimensions
(e.g. 2 PCs compared to 2000 genes), and when trying to build models with only
a few orthogonal data features. To visualize the structure in our dataset, 
implement the `plot_pca()` function to both run PCA on a provided count matrix,
but then to also plot the first two PCs of the dataset using a scatterplot,
where each dot represents a sample and is colored by its respective sample name.

*Hint*: pay close attention to the `prcomp` documentation for running PCA, and
ensure you're providing the count matrix in the correct orientation for running
PCA over the _samples_ in `plot_pca()`.
```{r}
#Display your PCA plot

```

## 4. Normalizing with counts per million

Now that we've explored our dataset prior to count normalization, we should 
normalize our data to see how things change. A simple and intuitive approach to
count normalization is _counts per million_: that is finding the number of counts 
for a given gene in a given sample for each million reads observed. It can be
easily calculated using this formula

$cpm = \frac{X_{i, j}}{\sum \limits_{i=1}^P X_{i, j}} \cdot 10^6$

Where $X$ is a $P \times N$ count matrix with genes as rows and samples as
columns.

### 4a. CPM matrix calculation

Implement the `normalize_by_cpm` function to normalize a raw count matrix given
the provided formula. Often we will want to work with log-transformed counts
as they are useful for stabilizing variances and possibly fitting assumed
statistical distributions for common tests. After calculating the CPM matrix,
ensure the returned matrix is `log2` transformed. 
```{r}
#Normalize your counts matrix (no output display expected)
```

### 4b. Visualizing CPM 

Visualize the effect of CPM normalization on the dataset by visualizing the
sample distributions, the relationship between variance and average expression
values, and plotting the samples along the first two PCs. Does the normalization
seem effective? What, if any, are the major differences observed between the
plots produced by raw counts?
```{r}
#Display 3 plots of your cpm matrix: Sample Distribution, Variance vs Mean, and PCA

```

## 5. Normalizing with DESeq2

DESeq2 is a common and prolific tool for performing differential expression
analysis (DEA) in RNAseq data. Within that process, DESeq2 implements its own
normalization procedure prior to performing DEA for more reliably identified
differential genes. You can read more about it [here](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#data-transformations-and-visualization).

### 5a. Variance stabilized counts with DESeq2

To normalize our data with DESeq2, implement the `deseq_normalize()` function
to extract the variance stabilized counts. Unlike CPM normalization, DESeq
directly considers the comparison of interest (in our case `timepoint`), so 
ensure you are both appropriately passing the sample metadata to the `deseq_normalize`
function as well as a formula describing the comparison of interest. 

*Hint*: the previously link DESeq2 documention explicity describes how to set up 
and calculate normalized values. For our case, we are interested in the counts 
generated using the `DESeq2::vst` function.
```{r}
#Normalize your DESeq2 data (no output display expected)
```

### 5b. Visualizing DESeq2 Normalizion

Visualize the effect of DESeq2 normalization on the dataset by visualizing the
sample distributions, the relationship between variance and average expression
values, and plotting the samples along the first two PCs. How does DESeq2
normalization compare to CPM? Raw values? Which method, if any, seems most
effective?
```{r}
#Display 3 plots of your normalized deseq data: Sample Distribution, Variance vs Mean, and PCA

```
